---
title: "Getting & Cleaning Data Project"
author: "antekai"
date: "Sunday, July 27, 2014"
output:
  html_document:
    keep_md: yes
---


Loading required packages.

```{r}
library(data.table)
library(reshape2)
```


###Getting data

```{r, eval=FALSE}
url <- "https://d396qusza40orc.cloudfront.net/getdata%2Fprojectfiles%2FUCI%20HAR%20Dataset.zip"
fileName <- "Dataset.zip"
if (!file.exists(fileName)) {
download.file(url, fileName, method="auto")}
unzip(fileName)
```

The archive put the files in a folder named `UCI HAR Dataset`. Setting this folder as the input path.

```{r}
pathIn <- file.path("UCI HAR Dataset")
```

`UCI HAR Dataset` contents:
```{r}
list.files(pathIn, recursive = TRUE)
```
See the [README.txt](https://github.com/antekai/GCDproject/uci har dataset/blob/master/readme.txt) for detailed information on the dataset.  
*For the purposes of this project, the files in the `Inertial Signals` folders are not used.*

###Reading data

Reading the subject files.

```{r}
dtSubjectTrain <- fread(file.path(pathIn,"train", "subject_train.txt"))
dtSubjectTest <- fread(file.path(pathIn, "test" , "subject_test.txt" ))
```

Reading the activity files

```{r}
dtActivityTrain <- fread(file.path(pathIn, "train", "Y_train.txt"))
dtActivityTest <- fread(file.path(pathIn, "test" , "Y_test.txt" ))
```

Reading the data files. 

```{r fileToDataTable}
fileToDataTable <- function (fileName) {
df <- read.table(fileName)
dt <- data.table(df)
}
dtTrain <- fileToDataTable(file.path(pathIn, "train", "X_train.txt"))
dtTest <- fileToDataTable(file.path(pathIn, "test" , "X_test.txt" ))
```


##Merge the training and the test sets


Concatenating the data tables.

```{r}
dtSubject <- rbind(dtSubjectTrain, dtSubjectTest)
setnames(dtSubject, "V1", "subject")
dtActivity <- rbind(dtActivityTrain, dtActivityTest)
setnames(dtActivity, "V1", "activityNum")
dt <- rbind(dtTrain, dtTest)
```

Merging columns.

```{r}
dtSubject <- cbind(dtSubject, dtActivity)
dt <- cbind(dtSubject, dt)
```

Setting key.

```{r}
setkey(dt, subject, activityNum)
```


##Extract only the mean and standard deviation

To find which variables in dt are measurements for the mean and standard deviation we read [features.txt](https://github.com/antekai/GCDproject/uci har dataset/blob/master/features.txt).

```{r}
dtFeatures <- fread(file.path(pathIn, "features.txt"))
setnames(dtFeatures, names(dtFeatures), c("featureNum", "featureName"))
```

Subsetting only measurements for the mean and standard deviation.

```{r}
dtFeatures <- dtFeatures[grepl("mean\\(\\)|std\\(\\)", featureName)]
```

Converting the column numbers to a vector of variable names matching columns

```{r}
dtFeatures$featureCode <- dtFeatures[, paste0("V", featureNum)]
head(dtFeatures)
dtFeatures$featureCode
```

Subsetting these variables using variable names.

```{r}
select <- c(key(dt), dtFeatures$featureCode)
dt <- dt[, select, with=FALSE]
```


##Use descriptive activity names


Check [activity_labels.txt](https://github.com/antekai/GCDproject/uci har dataset/blob/master/activity_labels.txt). This file will be used to add descriptive names to the activities.

```{r}
dtActivityNames <- fread(file.path(pathIn, "activity_labels.txt"))
setnames(dtActivityNames, names(dtActivityNames), c("activityNum", "activityName"))
```


##Label with descriptive activity names


Merging activity labels.

```{r}
dt <- merge(dt, dtActivityNames, by="activityNum", all.x=TRUE)
```

Adding `activityName` as a key.

```{r}
setkey(dt, subject, activityNum, activityName)
```

Melting the data table to reshape it from a short and wide format to a tall and narrow format.

```{r}
dt <- data.table(melt(dt, key(dt), variable.name="featureCode"))
```

Merging activity name.

```{r}
dt <- merge(dt, dtFeatures[, list(featureNum, featureCode, featureName)], by="featureCode", all.x=TRUE)
```

Creating a new variable `activity` that is equivalent to `activityName` as a factor class and a new variable `feature` that is equivalent to `featureName` as a factor class.

```{r}
dt$activity <- factor(dt$activityName)
dt$feature <- factor(dt$featureName)
```

Seperating features from `featureName` using the helper function `grepthis`.

```{r grepthis}
grepthis <- function (regex) {
  grepl(regex, dt$feature)
}
## Features with 2 categories
n <- 2
y <- matrix(seq(1, n), nrow=n)
x <- matrix(c(grepthis("^t"), grepthis("^f")), ncol=nrow(y))
dt$featDomain <- factor(x %*% y, labels=c("Time", "Freq"))
x <- matrix(c(grepthis("Acc"), grepthis("Gyro")), ncol=nrow(y))
dt$featInstrument <- factor(x %*% y, labels=c("Accelerometer", "Gyroscope"))
x <- matrix(c(grepthis("BodyAcc"), grepthis("GravityAcc")), ncol=nrow(y))
dt$featAcceleration <- factor(x %*% y, labels=c(NA, "Body", "Gravity"))
x <- matrix(c(grepthis("mean()"), grepthis("std()")), ncol=nrow(y))
dt$featVariable <- factor(x %*% y, labels=c("Mean", "SD"))
## Features with 1 category
dt$featJerk <- factor(grepthis("Jerk"), labels=c(NA, "Jerk"))
dt$featMagnitude <- factor(grepthis("Mag"), labels=c(NA, "Magnitude"))
## Features with 3 categories
n <- 3
y <- matrix(seq(1, n), nrow=n)
x <- matrix(c(grepthis("-X"), grepthis("-Y"), grepthis("-Z")), ncol=nrow(y))
dt$featAxis <- factor(x %*% y, labels=c(NA, "X", "Y", "Z"))
```

Checking to make sure all possible combinations of `feature` are accounted for by all possible combinations of the factor class variables.

```{r}
r1 <- nrow(dt[, .N, by=c("feature")])
r2 <- nrow(dt[, .N, by=c("featDomain", "featAcceleration", "featInstrument", "featJerk", "featMagnitude", "featVariable", "featAxis")])
r1 == r2
```


##Create a tidy data set


Creating a data set with the average of each variable for each activity and each subject.

```{r}
setkey(dt, subject, activity, featDomain, featAcceleration, featInstrument, featJerk, featMagnitude, featVariable, featAxis)
dtTidy <- dt[, list(count = .N, average = mean(value)), by=key(dt)]
```

Tidy data dimension
```{r,echo=F}
dim(dtTidy)
```

Saving tidy data to a comma-delimited csv file called `tidyData.csv`.
```{r}
ti <- file.path("tidyData.csv")
write.table(dtTidy, ti, quote = FALSE, sep = ",", row.names = FALSE)
```

